---
title: "Instacart_EDA_MBA"
author: "Tom Vanichachiva"
date: "March 4, 2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instacart Exploratory Data Analysis and Market Basket Analysis

In this analysis I will explore transactional Instacart data provided by Kaggle.

Kaggle is a data science and machine learning platform that allows users to compete and collaborate
on data science competitions.

The datasets in this competition were too large to push to GitHub. 
The datasets can be found here: https://www.kaggle.com/c/instacart-market-basket-analysis/data


## Loading Libraries
```{r message = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(arules)
library(cowplot)
library(scales)
library(arules)
library(arulesViz)
```

## Load in datasets
```{r }
aisles <- read.csv("data/aisles.csv")
department <- read.csv("data/departments.csv")
products_prior <- read.csv("data/order_products__prior.csv")
products_train <- read.csv("data/order_products__train.csv")
orders <- read.csv("data/orders.csv")
products <- read.csv("data/products.csv")
```

## Inspect the data

The Aisle dataset provides category information for products
The Category dataset also provides additional cateogry information for products
The Products Prior/Test datasets provide information on details of each transaction
The Orders dataset provides data on order metadata such as what time of day the order was placed
The Products dataset provides information on product metadata such as the name of the product

```{r}
glimpse(aisles) 
glimpse(department) 
glimpse(products_prior) 
glimpse(products_train) 
glimpse(orders) 
glimpse(products) 
```

## Convert character variables to factors

Converting a variable to a factor is a way for R to recognize a variable 
as a categorical variable. This will come in useful when we need to plot categorical variables.

```{r}
aisles <- aisles %>% 
            mutate(aisle = as.factor(aisle))

department <-  department %>% 
                mutate(department = as.factor(department))

orders <-  orders %>% 
            mutate(eval_set = as.factor(eval_set)
                   ,order_hour_of_day = as.numeric(order_hour_of_day)
                   ,order_dow = as.factor(order_dow)
                   ,order_hour_of_day = as.factor(order_hour_of_day))

products <- products %>% 
              mutate(product_name = as.factor(product_name))
```

## Top 10 Most Ordered Items

In both the prior/train datasets it looks like bananas and organic bananas are the most ordered items.

These plots will compare side by side to see how different the prior dataset is from the train dataset. 

```{r}
top10_products_prior <- products_prior %>% 
                        group_by(product_id) %>%
                        summarise(number_of_orders = n()) %>% 
                        top_n(10) %>%
                        left_join(products, by = "product_id") %>%
                        select(product_name, number_of_orders) %>%
                        arrange(desc(number_of_orders))

top10_products_train <- products_train %>% 
                        group_by(product_id) %>%
                        summarise(number_of_orders = n()) %>% 
                        top_n(10) %>%
                        left_join(products, by = "product_id") %>%
                        select(product_name, number_of_orders) %>%
                        arrange(desc(number_of_orders))

prod_bar_plot <- function(ds) {
  ds %>% 
    ggplot(aes(reorder(product_name, number_of_orders), number_of_orders, fill = product_name)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    labs(x = "Product Name", y = "Count") +
    theme_minimal()  +
    theme(legend.position = "none") 

} 

prod_prior_plt <- prod_bar_plot(top10_products_prior)
prod_train_plt <- prod_bar_plot(top10_products_train)


plot_grid(prod_prior_plt + labs(title = "Top 10 Items in the Prior Dataset")
          ,prod_train_plt + labs(title =  "Top 10 Items in the Train Dataset"))

```


## Average Instacart Basket Size - Train/Prior

The average basket size peaks around 5 items per transaction. This might signify that most people using Instacart are not doing all of their shopping on Instacart, rather they are using Instacart to pick up a few items. 

The distribution is positively skewed (right skewed). If these datasets were recreated in 2020, the distribution would probably be more negatively skewed (left skewed), i.e. the average basket size is probably larger.

```{r}

basket_size_prior <- products_prior %>% 
                      group_by(order_id) %>% 
                      summarise(basket_size = last(add_to_cart_order)) %>% 
                      ggplot(aes(basket_size)) +
                      geom_histogram(stat = "count", fill = "light blue", binwidth = 5) +
                      xlim(0,60) +
                      #scale_x_continuous(breaks = seq(from = 0, to = 60, by = 5)) +
                      theme_minimal()
                      
basket_size_train <- products_train %>% 
                      group_by(order_id) %>% 
                      summarise(basket_size = last(add_to_cart_order)) %>% 
                      ggplot(aes(basket_size)) +
                      geom_histogram(stat = "count", fill = "light green") +
                      xlim(0,60) +
                      #scale_x_continuous(breaks = seq(from = 0, to = 60, by = 5)) +
                      theme_minimal()
                      
plot_grid(basket_size_prior + labs(title = "Basket Size Distribution - Prior")
          ,basket_size_train + labs(title = "Basket Size Distribution - Train")
          ,ncol = 1)
```


## Busiest Day of the Week

The busiest day of the week seems to be the day 0 and 1. One can assume this is Saturday/Sunday since this is when people typically get their groceries. There is no way to definitively determine this.

```{r}
orders %>% 
  group_by(order_dow) %>% 
  tally() %>% 
  arrange(desc(n)) %>%
  ggplot(aes(reorder(order_dow, -n), n, fill = order_dow)) +
    geom_bar(stat = "identity") +
    scale_y_continuous(labels =  comma) +
  labs(x = "Day of the Week", y = "Count") +
    theme_minimal()
```

## Busiest Time of Day

The busiest time of day, i.e. when the most orders are placed, occur in the late morning to early afternoon.

Sending push notifications to Instacart users during this time frame would be the most optimal since this is when users typically place orders.

```{r}
orders %>% 
  group_by(order_hour_of_day) %>% 
  tally() %>% 
  arrange(desc(n)) %>%
  ggplot(aes(order_hour_of_day, n, fill = order_hour_of_day)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels =  comma) +
  labs(x = "Hour of the Day", y = "Count") +
  theme_minimal()
```

## Put Data into Basket Format

Putting data into a basket format means there is 1 record per order and all of the products on the order are also on the one record.

For example, in the original data set there could be 3 records for order 1. Each record is a different item. This step will take those 3 records for order 1 and combine it into 1 record.

This step is required for the Apriori Algorithim to read in the Instacart transaction data. 

```{r, results = FALSE}
train_basket <- products_train %>% 
                    left_join(products, by = "product_id") %>%  
                    select(order_id, product_name) 

#Write basket data to a transaction file then read it in
write.csv(train_basket, file = "data/train_basket.csv")
baskets <- read.transactions("data/train_basket.csv", format = "single", sep = ",", cols = c(2,3))
summary(baskets)
```

## Inspecting the First Five Baskets

In the first five basksets we can see produce items within each transaction.

```{r}
#Inspecting the first 5 baskets
inspect(baskets[1:5])
```

## Establishing Association Rules

```{r, results = 'hide'}
basket_rules <- apriori(baskets, parameter = list(support = 0.005, confidence = 0.30))
summary(basket_rules)

```

```{r}
summary(basket_rules)
inspect(basket_rules[1:5])
```


## Top Support and Confidence Rules

We can see that many transactions/rules show that bananas (organic/regular) are bought more regularly across the board.

For example, from the 1.3 million transactions it is very likely that when Organic Hass Avocados are bought so are Organic Bananas. In this dataset there was 2,420 instances of this occurring

```{r}
top_support <- sort(basket_rules, decreasing = TRUE, na.last = NA, by = "support")
inspect(head(top_support, 10))

top_confidence <- sort(basket_rules, decreasing = TRUE, na.last = NA, by = "confidence")
inspect(head(top_confidence, 10))

```

## Visualizing Association Rules

```{r}
plot(basket_rules)
inspectDT(basket_rules)
```

